hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-06-18 inst/manuscript/paper.Rmd
Editor comments (most)

Understanding the timing of key life history events is essential for effective fish conservation and management. Traditionally, predicting hatch and emergence timing in wild fish populations was challenging due to the reliance on average incubation temperature as a primary model parameter, which is often difficult to obtain in natural settings. Recent advancements have refined these models, enabling their application in wild environments using spawning dates and daily water temperature records. However, their broader use remains constrained by a lack of parameterizations for many species, with most applications focused on Salmonids. Here we introduce **hatchR**, a software ecosystem designed to predict hatch and emergence timing for a wide range of wild fishes, requiring only spawning date and wtaer temperature data. **hatchR** offers users access to established phenological models and the flexibility to incorporate custom parameterizations using external datasets. The software is available in two formats: an open-source R package for advanced customization and an HTML-based graphical user interface for those unfamiliar with scripting. To illustrate its utility, we present two case studies demonstrating its application in research and management. By expanding access to predictive modeling tools, **hatchR** has the potential to advance studies of fish early life history and support conservation efforts across diverse species.

@@

The effective value approach has since been widely applied to Salmonids, for which aquaculture-derived parameterizations are readily available. For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019. Despite its growing adoption, applications of the effective value model remain largely confined to Salmonids, likely due to the availability of existing parameterizations and the commercial, recreational, and cultural importance of these species.

To extend these modeling capabilities beyond Salmonids and facilitate broader applications, we developed **hatchR**, a software ecosystem designed to predict hatch and emergence timing for wild fish populations. **hatchR** enables users to input standard raw or summarized water temperature datasets commonly collected in field settings, conduct basic data validation, and apply built-in parameterizations such as those from @beacham1990 or @sparks2017. Users can also develop custom models using their own or published temperature and phenological data within the effective value framework for application to unparameterized species or populations.

@@

**hatchR** is designed primarily as a tool for predicting early life history phenology in wild fishes. To maintain focus on this core function, we provide minimal built-in data validation and visualization tools, as users are expected to understand and check their own data. Given the diversity of potential data types, it is impractical to implement comprehensive validation checks. However, we include basic data-checking and summarization functions (`check_continuous()`, `summarize_temp()`) and limited built-in visualization capabilities (`plot_check_temp()`, `plot_phenology()`), the process for which is highlighted in **Box XXX**. Intuitive functions are provided for users to apply models—either existing models from the literature using the `model_select()` function or fitting custom functions from data using the `fit_model()` function. Users can then apply these models to water temperature data (*e.g.,* from an in-stream temperature logger) to predict when hatching phenology will occur. This is accomplished with the `predict_phenology()` function. The R package provides user-friendly example articles that build in complexity for teaching users the functionality and flexibility of each aspect of the work flow as well as customization for plotting model outputs. In contrast, the Shiny application includes a default output plot and an option to download results for external visualizations. For a high-level overview of **hatchR**'s applications, see Figure 1. Additional details on key functions and workflows, particularly for automating phenology predictions across multiple variables, are available in articles hosted on the software's webpage.

@@

This framework is the foundation for phenological models in **hatchR**. The package includes a predefined `model_table` containing established parameterizations, including those from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979). While `model_table` incorporated more complex models from @beacham1990, users can also fit custom models using the `fit_model()` function. This flexibility allows for the incorporation of new parameterizations as they are developed, expanding the utility of **hatchR** beyond Salmonids.

@@

**hatchR** assumes input data consists of at least two required columns: a date column indicating the date (and optionally time in the case of fiedl) of each temperature measurement, and a temperature column providing the corresponding temperature measurement (in °C). Other columns may be present, but column names should not include spaces. Data should follow the format outlined in Table 1.

@@

**hatchR** will function with missing values, but gaps in the dataset may affect predictions. Since the package does not automatically handle missing data, users must check for gaps or errors before running analyses with the provided check functions (see section *Checking Data*). **hatchR** supports temperatures as low as 0 °C, though such values yield extremely small effective values, potentially extending hatch or emergence timing to a year or more. Users should critically assess whether such data align with biological expectations.

@@

**hatchR** is designed to analyze daily average temperatures. While high-frequency data (*e.g.*, from HOBO loggers) can be used, it must be summarized into daily averages. **hatchR** provides built-in functionality for this summarization in R but requires pre-summarized data for use in the Shiny app.

To help users identify potential issues, **hatchR** includes basic data checking functions that highlight outliers or missing values both visually and programatically. These checks ensure data integrity before model application.

Here, we demonstrate the utility of three functions--`summarize_temp()`, `plot_check_temp()`, and `check_continuous()`--using a simulated year-long dataset (`year_sim`). This dataset contains temperature readings taken every thirty minutes, and its structure (dimensions and first six rows) is shown below:

We recommend using `plot_check_temp()` to visually inspect imported data for outliers or unusual values (Figure 3). This function plots the temperature data against time, allowing users to set custom thresholds for minimum and maximum temperatures. In this example, we set the minimum and maximum thresholds to 0 and 25 °C, respectively.

# **hatchR** Workflow

Users can select from published model parameterizations contained in `model_table`, or generate custom model parameterizations using `fit_model()`, in both the R and Shiny deployments of **hatchR**. The models in `model_table` are included because their parameterizations are well-documented in the literature, though they are currently limited to Pacific Salmon and Bull Trout (see @quinn2018 pg. 183, for additional Salmonid models). To ensure reliability, we restrict `model_table` to well vetted models with experimental ranges spanning 2-17 °C, instead requiring users to carefully asses whether their own parameterized models are appropriate for the temperature ranges in their datasets.

## Understanding your results

The output from `predict_phenology()` (in this example, our named object `WI_hatch`) contains multiple elements in a list, which can be accessed using the `$` operator. Each component provides different insight into the predicted phenology:

## Plotting phenology

**hatchR** includes a built in function, `plot_phenology()`, for visualizing phenology predictions (Figure 4). This function generates plots with three specific components: 1) the temperature regime over the prediction period, 2) the cumulative sum of effective values, and 3) the effective value for each day within the prediction span. Be default, `plot_phenology()` produces a comprehensive figure that includes all three elements with corresponding labels and titles. However, users can customize the output to focus on specific aspects of interest, allowing for tailored visual representations of their results. This function provides a quick and effective way to interpret model outputs, facilitating comparisons between temperature regimes or species-specific phenological responses.

```{r, fig.cap="Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` dataset."}
plot_phenology(WI_hatch)
```

@@

To expand the applicability of the effective value approach beyond Salmonids, **hatchR** includes a `fit_model()` function, which only requires species' development follows a power law relationship with temperature. The function takes two input vectors: average incubation temperature (°C) and the number of days to a given phenological event. A model is then fit to the data using `stats::nls()`, which performs nonlinear least squares regression to estimate the parameters *log~e~a* and *b*. Because the optimization process in `nls()` is sensitive to initial parameter values, `fit_model()` first fits a linear model to the log-transformed data to provide initial parameter estimates, which are then used to

This approach allows users to generate models tailored to any species of fish, or even different populations within species, provided they have experimental or field data linking development to temperature. However, users should be mindful of several factors, such as extrapolation risks (models may not generalize beyond the temperature range for which they were parameterized) and species-specific variation (genetic differences among populations may affect developmental responses). Future expansions of **hatchR** could incorporate additional vetted parameterizations for other taxa, such as non-Salmonid fishes, amphibians, or invertebrates, provided sufficient validation in the literature.

To demonstrate how `fit_model()` may be used to create custom parameterizations for species beyond the Salmonids included in `model_table`, we provide parameterizations for three warm-water species: Smallmouth Bass (*Micropterus dolomieu*) [@webster1948], Channel Catfish (*Ictalurus punctatus*) [@small2001], and Lake Sturgeon (*Acipenser fulvescens*) [@smith2005]. These species were selected due to their common use in aquaculture and sport fisheries. For concision, we present parameterization for Smallmouth Bass below, while the full implementation details for all species are available in the `paper.Rmd` on the GitHub project repository (<https://github.com/bmait101/hatchR>).

Finally, we summarize the outputs across species to visualize hatch timing and total developmental duration (Figure 5). This example highlights the flexibility of **hatchR** for accommodating diverse fish species and environmental conditions, making it a valuable tool for researchers and managers working outside of Salmonid systems.

@@

Once uploaded, **hatchR** automatically generates a visual data check using `plot_check_temp()`. After confirming data integrity, users navigate to the `Model Phenology` window. For this case study, we use the pre-loaded Bull Trout parameterization from @austin2019, selecting the `Existing` model option via dropdown menus. The user then may chooses multiple spawn dates using an interactive calendar. Here, we focus on September 30th (in the 2014 spawn year) as outlined in our management scenario.

@@

**hatchR** is a software ecosystem that bridges the analytical gap in predicting developmental phenology for wild fishes. It establishes a formal framework for applying effective value models from user-provided parameterizations. The software is available in two formats: 1) A fully customizable R package, ideal for complex and repetitive analyses and 2) a graphical user interface for ease of use, designed for tasks that may only need to be run once or a few times.

Both versions allow users to import data, perform basic data checks, perform basic visualizations, and apply either pre-existing Salmonid model parameterizations or generate custom models specific to other species or populations. To support users at various levels of expertise, we provide extensive documentation on the **hatchR** website (<https://bmait101.github.io/hatchR/>), covering basic and advanced applications.

@@

Second, developmental timing occurs as a distribution, not a fixed point. While **hatchR** provides point estimates for developmental phenology, fish spawning and development within populations occur as distributions rather than single events [@mason1976]. We encourage users to predict phenology using early, peak, and late thresholds (*e.g.*, 5th, 50th, and 95th percentiles) or incorporate real or modeled distributions to capture variation.

@@

>>>>

Understanding the timing of key life history events is essential for effective fish conservation and management. Traditionally, predicting hatch and emergence timing in wild fish populations was challenging due to the reliance on average incubation temperature as a primary model parameter, which is often difficult to obtain in natural settings. Recent advancements have refined these models, enabling their application in wild environments using spawning dates and daily water temperature records. However, their broader use remains constrained by a lack of parameterizations for many species, with most applications focused on salmonids. Here we introduce **hatchR**, a software ecosystem designed to predict hatch and emergence timing for a wide range of wild fishes, requiring only spawning date and wtaer temperature data. **hatchR** offers users access to established phenological models and the flexibility to incorporate custom parameterizations using external datasets. The software is available in two formats: an open-source R package for advanced customization and an HTML-based graphical user interface for those unfamiliar with scripting. To illustrate its utility, we present two case studies demonstrating its application in research and management. By expanding access to predictive modeling tools, **hatchR** has the potential to advance studies of fish early life history and support conservation efforts across diverse species.

@@

The effective value approach has since been widely applied to salmonids, for which aquaculture-derived parameterizations are readily available. For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019. Despite its growing adoption, applications of the effective value model remain largely confined to Salmonids, likely due to the availability of existing parameterizations and the commercial, recreational, and cultural importance of these species.

To extend these modeling capabilities beyond salmonids and facilitate broader applications, we developed **hatchR**, a software ecosystem designed to predict hatch and emergence timing for wild fish populations. **hatchR** enables users to input standard raw or summarized water temperature datasets commonly collected in field settings, conduct basic data validation, and apply built-in parameterizations such as those from @beacham1990 or @sparks2017. Users can also develop custom models using their own or published temperature and phenological data within the effective value framework for application to unparameterized species or populations.

@@

The **hatchR** softeware is designed primarily as a tool for predicting early life history phenology in wild fishes. To maintain focus on this core function, we provide minimal built-in data validation and visualization tools, as users are expected to understand and check their own data. Given the diversity of potential data types, it is impractical to implement comprehensive validation checks. However, we include basic data-checking and summarization functions (`check_continuous()`, `summarize_temp()`) and limited built-in visualization capabilities (`plot_check_temp()`, `plot_phenology()`), the process for which is highlighted in Box 1. Intuitive functions are provided for users to apply models—either existing models from the literature using the `model_select()` function (Boxes 2,3) or fitting custom functions from data using the `fit_model()` function (Box 4). Users can then apply these models to water temperature data (*e.g.,* from an in-stream temperature logger) to predict when developmental phenology will occur using the `predict_phenology()` function (Box 2). Alternatively, if users had known or observed hatch or emergence dates, they could use `predict_spawn()` to backcalculate spawn timing using the appropriate corresponding developmental models. Finally, users can either use model output (Box 3) for custom graphics or generate simple plots using the `plot_phenology()` function which can use the output object generated from `predict_phenology()` or `predict_spawn()` (Box 5).

The R package provides user-friendly example articles that build in complexity for teaching users the functionality and flexibility of each aspect of the work flow as well as customization for plotting model outputs. In contrast, the Shiny application includes a default output plot and an option to download results for external visualizations. For a high-level overview of **hatchR**'s applications, see Figure 1. Additional details on key functions and workflows, particularly for automating phenology predictions across multiple variables, are available in articles hosted on the software's webpage.

@@

This framework is the foundation for phenological models in **hatchR**. The package includes a predefined `model_table` containing established parameterizations, including those from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979). While `model_table` incorporated more complex models from @beacham1990, users can also fit custom models using the `fit_model()` function. This flexibility allows for the incorporation of new parameterizations as they are developed, expanding the utility of **hatchR** beyond salmonids.

@@

The **hatchR** software assumes input data consists of at least two required columns: a date column indicating the date (and optionally time in the case of field) of each temperature measurement, and a temperature column providing the corresponding temperature measurement (in °C). Other columns may be present, but column names should not include spaces. Data should follow the format outlined in Table 1.

@@

The **hatchR** software will function with missing values, but gaps in the dataset may affect predictions. Since the package does not automatically handle missing data, users must check for gaps or errors before running analyses with the provided check functions (see section *Checking Data*). The **hatchR** software supports temperatures as low as 0 °C, though such values yield extremely small effective values, potentially extending hatch or emergence timing to a year or more. Users should critically assess whether such data align with biological expectations.

@@

The **hatchR** software is designed to analyze daily average temperatures. While high-frequency data (*e.g.*, from HOBO loggers) can be used, it must be summarized into daily averages. The **hatchR** software provides built-in functionality for this summarization in R but requires pre-summarized data for use in the Shiny app.

To help users identify potential issues, **hatchR** includes basic data checking functions that highlight outliers or missing values both visually and programatically. These checks ensure data integrity before model application and are demonstrated in Box 1. Additionally, `plot_check_temp()` and `check_continuous()` are automatically run when data is loaded into the Shiny app and presented in the `Import Data` window.

Users can select from published model parameterizations contained in `model_table` (Box 2), or generate custom model parameterizations using `fit_model()` (Box 4) in both the R and Shiny deployments of **hatchR**. The models in `model_table` are included because their parameterizations are well-documented in the literature, though they are currently limited to Pacific salmon and Bull Trout (see @quinn2018 pg. 183, for additional salmonid models). To ensure reliability, we restrict `model_table` to well vetted models with experimental ranges spanning 2-17 °C, instead requiring users to carefully asses whether their own parameterized models are appropriate for the temperature ranges in their datasets.

@@

To expand the applicability of the effective value approach beyond salmonids, **hatchR** includes a `fit_model()` function, which only requires species' development follows a power law relationship with temperature (Box 4). The function takes two input vectors: average incubation temperature (°C) and the number of days to a given phenological event. A model is then fit to the data using `stats::nls()`, which performs nonlinear least squares regression to estimate the parameters *log~e~a* and *b*. Because the optimization process in `nls()` is sensitive to initial parameter values, `fit_model()` first fits a linear model to the log-transformed data to provide initial parameter estimates, which are then used to

This approach allows users to generate models tailored to any species of fish, or even different populations within species, provided they have experimental or field data linking development to temperature. However, users should be mindful of several factors, such as extrapolation risks (models may not generalize beyond the temperature range for which they were parameterized) and species-specific variation (genetic differences among populations may affect developmental responses). Future expansions of **hatchR** could incorporate additional vetted parameterizations for other taxa, such as non-salmonid fishes, amphibians, or invertebrates, provided sufficient validation in the literature.

@@

Once uploaded, **hatchR** automatically generates a visual data check using `plot_check_temp()`. After confirming data integrity, users navigate to the `Model Phenology` window. For this case study, we use the pre-loaded Bull Trout parameterization from @austin2019, selecting the `Existing` model option via dropdown menus. The user then may chooses a single or multiple spawn dates using an interactive calendar. Here, we focus on September 30th (in the 2014 spawn year) as outlined in our management scenario.

@@

The **hatchR** software ecosystem bridges the analytical gap in predicting developmental phenology for wild fishes. It establishes a formal framework for applying effective value models from user-provided parameterizations. The software is available in two formats: 1) A fully customizable R package, ideal for complex and repetitive analyses and 2) a graphical user interface for ease of use, designed for tasks that may only need to be run once or a few times.

Both versions allow users to import data, perform basic data checks, perform basic visualizations, and apply either pre-existing salmonid model parameterizations or generate custom models specific to other species or populations. To support users at various levels of expertise, we provide extensive documentation on the **hatchR** website (<https://bmait101.github.io/hatchR/>), covering basic and advanced applications.

@@

Second, developmental timing occurs as a distribution, not a fixed point. While **hatchR** provides point estimates for developmental phenology, fish spawning and development within populations occur as distributions rather than single events [@mason1976]. We encourage users to predict phenology using early, peak, and late thresholds (*e.g.*, 5th, 50th, and 95th percentiles) or incorporate real or modeled distributions to capture variation. We outline an approach for predicting phenology from a modeled distribution in our "Advanced plotting" article on the **hatchR** website (<https://bmait101.github.io/hatchR/articles/Advanced_plotting.html>).

@@

# Box 1: Checking and summarizing your data

To demonstrate the process of checking your data we show the workflow for the functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long dataset (`year_sim`). This dataset contains temperature readings taken every thirty minutes, and its structure (dimensions and first six rows) is shown below:

We recommend using `plot_check_temp()` to visually inspect imported data for outliers or unusual values. This function plots the temperature data against time, allowing users to set custom thresholds for minimum and maximum temperatures. In this example, we set the minimum and maximum thresholds to 0 and 25 °C, respectively.

# Box 2: Model selection and prediction

# Box 3: Understanding model output

The output from `predict_phenology()` (in this example, our named object `WI_hatch` generated in Box 2) contains multiple elements in a list, which can be accessed using the `$` operator. Each component provides different insight into the predicted phenology:

# Box 4: Generating custom models

To demonstrate how `fit_model()` may be used to create custom parameterizations for species beyond the salmonids included in `model_table`, we provide parameterizations for three warm-water species: Smallmouth Bass (*Micropterus dolomieu*) [@webster1948], Channel Catfish (*Ictalurus punctatus*) [@small2001], and Lake Sturgeon (*Acipenser fulvescens*) [@smith2005]. These species were selected due to their common use in aquaculture and sport fisheries. For concision, we present parameterization for Smallmouth Bass below, while the full implementation details for all species are available in the `paper.Rmd` on the GitHub project repository (<https://github.com/bmait101/hatchR>).

Finally, we summarize the outputs across species to visualize hatch timing and total developmental duration. This example highlights the flexibility of **hatchR** for accommodating diverse fish species and environmental conditions, making it a valuable tool for researchers and managers working outside of salmonid systems.

# Box 5: Plotting phenology

The **hatchR software** includes a built in function, `plot_phenology()`, for visualizing phenology predictions. This function generates plots with three specific components: 1) the temperature regime over the prediction period, 2) the cumulative sum of effective values, and 3) the effective value for each day within the prediction span. By default, `plot_phenology()` produces a comprehensive figure that includes all three elements with corresponding labels and titles. However, users can customize the output to focus on specific aspects of interest, allowing for tailored visual representations of their results. This function provides a quick and effective way to interpret model outputs, facilitating comparisons between temperature regimes or species-specific phenological responses.

```{r, fig.cap="Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` dataset."}
plot_phenology(WI_hatch)

```
